---
title: "California housing: reading the glass box with tslr's plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{California housing: reading the glass box with tslr's plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 9, fig.height = 5, dpi = 110, fig.retina = 1,
  warning = FALSE, message = FALSE
)

# Locate the California housing CSV. Override with the TSL_CALIFORNIA_CSV
# environment variable; otherwise search upward from the working directory for
# `data/housing_full.csv` (present in the source repository).
find_california <- function() {
  env <- Sys.getenv("TSL_CALIFORNIA_CSV", "")
  if (nzchar(env) && file.exists(env)) return(env)
  dir <- normalizePath(getwd(), mustWork = FALSE)
  repeat {
    cand <- file.path(dir, "data", "housing_full.csv")
    if (file.exists(cand)) return(cand)
    parent <- dirname(dir)
    if (identical(parent, dir)) break
    dir <- parent
  }
  ""
}

csv_path <- find_california()
have_data <- nzchar(csv_path)

# The figure chunks need the dataset; gate them in one place so a standalone
# build (no data file) degrades to prose only.
knitr::opts_chunk$set(eval = have_data)
```

```{r note, eval = TRUE, echo = FALSE, results = "asis"}
if (!have_data) {
  cat("> **Note.** This vignette renders its figures from the California housing",
      "dataset (`data/housing_full.csv`). The data file was not found, so the",
      "figures are skipped here; the prose and code still show how each one is",
      "built. Point `TSL_CALIFORNIA_CSV` at the CSV to render them.\n")
}
```

TSL is a *glass box*: a fitted model is a sum of boosting **stages**, and every
stage factorises into per-feature univariate pieces. Because of that separability
`tslr` can reconstruct -- exactly, not by sampling -- partial dependence, spatial
effects, tilt, feature importance, and per-point explanations straight from the
fitted model. This vignette walks through those built-in plotting functions on the
classic California housing data; each renders with [ggplot2](https://ggplot2.tidyverse.org/)
in a flat, hairline style and returns an ordinary `ggplot` object you can extend or
`ggsave()`.

```{r libs}
library(tslr)
library(ggplot2)
```

## Data and model

The data are the eight standard California housing predictors with the median
house value (in units of \$100,000) as the response. Longitude and latitude let
us draw the model's spatial structure on the California map.

```{r data}
raw <- as.matrix(utils::read.csv(csv_path, header = FALSE))
feature_names <- c("MedInc", "HouseAge", "AveRooms", "AveBedrms",
                   "Population", "AveOccup", "Latitude", "Longitude")
X <- raw[, 1:8]
colnames(X) <- feature_names
y <- raw[, 9]                      # median house value, in $100k
```

```{r fit}
model <- tsl(X, y, epochs = 5L, n_trees = 12L, n_iter = 25L, split_try = 16L,
             colsample_bytree = 1.0, seed = 0L, verbosity = 0L)
pred <- predict(model, X)
cat(sprintf("stages = %d   train RMSE = %.3f ($100k)\n",
            length(tsl_components(model)), sqrt(mean((pred - y)^2))))
```

`tsl()` keeps a copy of the training matrix on the fitted object, so the plotting
functions below need no further data -- they marginalise over it automatically.
Every function also accepts an explicit `X` if you want to integrate over a
different background.

## Partial dependence

`plot_first_order_pd()` shows, per stage, the positive branch (orange), the
negative branch (blue), and their net (the solid line) -- the marginal effect of
each feature with everything else integrated over the data.

```{r pd1d, fig.width = 11, fig.height = 9}
plot_first_order_pd(model, features = c("Latitude", "Longitude", "MedInc"),
                    show_data_density = TRUE)
```

`pd_difference_plot()` adds the backbone overlay and makes TSL's signature visible:
a large positive branch and a large negative branch that mostly cancel, leaving a
small net effect.

```{r pddiff, fig.width = 11, fig.height = 9}
pd_difference_plot(model, features = c("Latitude", "Longitude", "MedInc"))
```

## Spatial structure on the California map

`plot_2d_backbone()` draws, per stage, the backbone product `b_lon * b_lat`
(where the stage is "switched on") above the signed 2D partial dependence (the
location effect on price).

```{r backbone, fig.width = 12, fig.height = 8}
plot_2d_backbone(model, feature_x = "Longitude", feature_y = "Latitude",
                 stages = 1:3)
```

Every plot is an ordinary ggplot built from a tidy data frame, so you can drop to
the **compute layer** ([`tsl_pd_2d()`], [`tsl_backbone_2d()`], ...) and build a
custom figure. Here we put the stage-1 signed surface on the actual California
map, overlaying a faint scatter of the observations that traces the coast and
Central Valley.

```{r map, fig.width = 7, fig.height = 6}
surf <- tsl_pd_2d(model, feature_x = "Longitude", feature_y = "Latitude",
                  grid_points = 90L, stages = 1L)
pts <- as.data.frame(X[sample(nrow(X), 3000), c("Longitude", "Latitude")])
names(pts) <- c("lon", "lat")

ggplot(surf, aes(x, y, fill = value)) +
  geom_tile() +
  geom_point(data = pts, aes(lon, lat), inherit.aes = FALSE,
             colour = "#18181B", alpha = 0.05, size = 0.12) +
  scale_fill_tsl_diverging(name = "2D PD",
                           limits = max(abs(surf$value)) * c(-1, 1)) +
  coord_quickmap(expand = FALSE) +
  labs(title = "Stage 1 location effect", x = "Longitude", y = "Latitude") +
  theme_flat()
```

The tilt -- the directional part of each stage -- has its own 2D view.

```{r tilt2d, fig.width = 12, fig.height = 4.6}
plot_2d_tilt(model, feature_x = "Longitude", feature_y = "Latitude",
             stages = 1:3)
```

## Tilt curves and diagnostics

`plot_tilt_1d()` draws the per-feature tilt `d_j(x)` by stage; stage 1 carries
almost no tilt (it is the magnitude stage) and later stages add the signed
corrections.

```{r tilt1d, fig.width = 11, fig.height = 9}
plot_tilt_1d(model, features = c("Latitude", "Longitude", "MedInc"))
```

`plot_tilt_diagnostics()` expands one feature into four views -- the raw tilt
squashed by `tanh`, the same weighted by the backbone, and both again after
centring -- so you can see where a tilt actually moves the prediction.

```{r tiltdiag, fig.width = 12, fig.height = 7}
plot_tilt_diagnostics(model, features = "Latitude")
```

## Feature importance

`plot_feature_importance()` is a six-panel dashboard: per-stage backbone and tilt
heatmaps, the stage weights, and the aggregated global bars (this one uses
**patchwork** to compose the panels).

```{r importance, fig.width = 14, fig.height = 8}
plot_feature_importance(model)
```

## Local explanations

`plot_local_interpretation()` decomposes individual predictions. A coastal house
and an inland desert house attribute their very different values to different
stages, backbone shares, and tilts.

```{r local, fig.width = 13, fig.height = 7}
coastal <- which.min(abs(X[, "Longitude"] - (-118.24)) +
                       abs(X[, "Latitude"] - 34.05))   # Los Angeles
desert  <- which.min(abs(X[, "Longitude"] - (-116.0)) +
                       abs(X[, "Latitude"] - 34.7))     # Mojave

plot_local_interpretation(model, list(X[coastal, ], X[desert, ]),
                          titles = c("Coastal", "Desert"))
```

## ICE curves

Finally, `plot_ice()` shows individual conditional expectation curves -- one per
sampled house -- with the partial-dependence mean overlaid.

```{r ice, fig.width = 7, fig.height = 4.5}
plot_ice(model, feature = "MedInc", n_ice = 60L)
```

Every figure here was produced by a single `tslr` function reading the fitted
model. Because TSL is separable, all of them -- partial dependence, spatial
structure, tilt, importance, and local attributions -- are exact read-outs of the
glass box rather than post-hoc approximations. Each returns a `ggplot` (or
patchwork) object; reach for the `tsl_*()` compute functions when you want the
underlying numbers or a bespoke figure.
